Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-03-22 paperDistance/paper.tex
small tweaks, more tomorrow

In this paper we use the denoised, observed \cmd\ as a prior to infer more precise parallax pdfs from \gaia\ measurements.
For a toy example to show how this method works, instead of using the \cmd\ as our prior to get more precise measurements of parallax, we use a straight line $y = mx + b$ with some gaussian thickness $\Delta$ as our prior to infer some posterior belief over $y_{true,n}$ from noisy $y_n$ data.
Here the thickness of the true distribution mimics the true thickness of the \cmd\ due to age, mass and metallicity.
To set up the problem we have the true underlying distribution $y_{true, n} = m_{true}\,x_n + b_{true} + \Delta_n$ where $\Delta_n$ is drawn from $\mathcal{N}(\Delta_n \given 0, t^2)$, where $\mathcal{N}$ represents a Gaussian function, here with mean $0$ and variance $t^2$.
We also have some noisy data $y_n$ with measurement uncertainty $\sigma_n$.
To generate a measurement uncertainty $\sigma_n$, we draw from a uniform distribution a value $z_n$, and then stretch the value to get some more significant outliers: $\sigma_n = (2z_n)^3$.
From this we generate our noisy data $y_n = y_{true,n} + \mathcal{N}(\mathrm{noise} \given 0, \sigma_n^2)$.
So we have a thick true distribution, and some noisy draws from this distribution.

First we determine the underlying, true distribution that the noisy data $y_n$ are drawn from.
This denoised distribution is our prior.
We then use this pior to infer a posterior belief about the true values, $p(y_{true,n} \given y_n, \tilde{\sigma}_n^2)$.
When the true values are inferred in this way, the associated variance $\tilde{\sigma}_n^2$ is the convolution of the true distribution variance $t^2$ and the measurement variance $\sigma_n^2$.
Figure \ref{fig:line}, left panel, shows 1024 realizations of the noisy values $y_n$ and their associated measurement uncertainties $\sigma_n$ as the grey points. Here we highlight 5 random points in black to highlight the method, but the same method is being applied to all the data.
We denoise these points using marginalized maximum likelihood estimation (MML), which marginalizes over the true data values (we never need to know them) and maximizes the likelihood of the data. Here our likelihood for each individual datum is $p(y_n \given m, b, t^2, \sigma_n^2) = \mathcal{N}(y_n \given m\,x_n + b, t^2 + \sigma_n^2)$ and the likelihood of the full data set is $\prod_n\, p(y_n \given m, b, t^2, \sigma_n^2)$. Maximizing this likelihood is the same as minimizing the $\chi^2 = \sum_n \frac{[y_n - [m\,x_n + b]]^2}{[\sigma_n^2 + t_n^2]} + \sum_n\,ln(2\pi\,[\sigma_n^2 + t^2])$. This MML gives us the best fit parameters $\hat{m}$, $\hat{b}$, and $\hat{t}$. In Figure \ref{fig:line}, center panel, the true functions $y_{true} = m_{true}\,x + b_{true} \pm t_{true}$ are shown as the red lines, and the best fit MML functions $\hat{y}_{true} = \hat{m}\, x + \hat{b} \pm \hat{t}$ are shown as the blue lines. Now that we have an estimate of the true underlying distribution $\hat{y}_{true}$, we use this as a prior to infer a posterior belief of the true values for each datum $p(y_{true,n} \given y_n, \tilde{\sigma}_n^2)$

So using Bayes theorem
\begin{eqnarray}
p(y_{true,n} \given y_n, \tilde{\sigma}_n) &=& p(y_n \given y_{true,n}, \sigma_n)\,p(y_{true,n}) \\
p(y_n \given y_{true,n}, \sigma_n) &=& \mathcal{N}(y_n \given y_{true,n}, \sigma_n) \\
p(y_{true,n}) &=& \mathcal{N}(y_{true,n} \given \hat{m}\, x_n + \hat{b}, \hat{t})
\label{eq:toyBayes}
\end{eqnarray}

Since everything is Gaussian, this can be done analytically by multiplying Gaussians. The blue points in Figure \ref{fig:line}, center panel, represent the posterior belief of the true value $p(y_{true,n} \given y_n, \tilde{\sigma}_n)$. The black points represent the original, noisy data $y_n \pm \sigma_n$, the same as those highlighted in panel A, and the red points are the true values $y_{true, n}$. Finally, Figure \ref{fig:line}, right panel, shows the full set of posterior beliefs for each $y_n$. The points are colored by their measurement variance $\sigma_n^2$ to show that noisier measurements are more influenced by the prior and get pulled closer to the center of the MML distribution, our prior, while more certain measurements remain closer to their measured values $y_n$.

>>>>

In this paper we use the denoised, observed \cmd\ as a prior to infer more precise parallax pdfs from \gaia\ measurements.
Here we will demonstrate this method with a simple toy mode. Instead of using the \cmd\ as our prior to get more precise measurements of parallax, we use a straight line $y = mx + b$ with some gaussian thickness $t$ as our prior. We use this prior to infer some posterior belief over $y_{\true,n}$ from noisy $y_n$ data.
Here the thickness of the true distribution mimics the true thickness of the \cmd\ due to age, mass and metallicity.
To set up the problem we have the true underlying distribution $y_{\true, n} = m_{\true}\,x_n + b_{\true} + \Delta_n$ where $\Delta_n$ is drawn from $\mathcal{N}(\Delta_n \given 0, t^2)$, where $\mathcal{N}$ represents a Gaussian function, here with mean $0$ and variance $t^2$.
We also have some noisy data $y_n$ with measurement uncertainty $\sigma_n$.
To generate a measurement uncertainty $\sigma_n$, we draw from a uniform distribution a value $z_n$, and then stretch the value to get some more significant outliers: $\sigma_n = (2z_n)^3$.
From this we generate our noisy data $y_n = y_{\true,n} + \epsilon$ where $\epsilon$ is drawn from $\mathcal{N}(\epsilon \given 0, \sigma_n^2)$.
So we have a thick true distribution, and some noisy draws from this distribution.

First we determine the underlying, true distribution that the noisy data $y_n$ are drawn from.
This denoised distribution is our prior.
We then use this pior to infer a posterior belief about the true values, $p(y_{\true,n} \given y_n, \tilde{\sigma}_n^2)$.
When the true values are inferred in this way, the associated variance $\tilde{\sigma}_n^2$ is the convolution of the true distribution variance $t^2$ and the measurement variance $\sigma_n^2$.
Figure \ref{fig:line}, left panel, shows 1024 realizations of the noisy values $y_n$ and their associated measurement uncertainties $\sigma_n$ as the grey points. Here we highlight 5 random points in black to highlight the method, but the same method is being applied to all the data.
We denoise these points using marginalized maximum likelihood estimation (MML), which marginalizes over the true data values (we never need to know them) and maximizes the likelihood of the data. Here our likelihood for each individual datum is $p(y_n \given m, b, t^2, \sigma_n^2) = \mathcal{N}(y_n \given m\,x_n + b, t^2 + \sigma_n^2)$ and the likelihood of the full data set is $\prod_n\, p(y_n \given m, b, t^2, \sigma_n^2)$. Maximizing this likelihood is the same as minimizing the $\chi^2 = \sum_n \frac{[y_n - [m\,x_n + b]]^2}{[\sigma_n^2 + t_n^2]} + \sum_n\,ln(2\pi\,[\sigma_n^2 + t^2])$. This MML gives us the best fit parameters $\hat{m}$, $\hat{b}$, and $\hat{t}$. In Figure \ref{fig:line}, center panel, the true functions $y_{\true} = m_{\true}\,x + b_{\true} \pm t_{\true}$ are shown as the red lines, and the best fit MML functions $\hat{y}_{\true} = \hat{m}\, x + \hat{b} \pm \hat{t}$ are shown as the blue lines. Now that we have an estimate of the true underlying distribution $\hat{y}_{\true}$, we use this as a prior to infer a posterior belief of the true values for each datum $p(y_{\true,n} \given y_n, \tilde{\sigma}_n^2)$. Here is a good point to step back and acknowledge that our method is odd/strange/different/new (which maybe you're also currently thinking but can't quite express why yet). We are using the set of data to determine the denoised distribution, and then using that denoised distribution to make inferences on the data. We are technically using the data twice! If we were going to be fully, technically correct, we should denoise the distribution, but leave one datum out; the one datum we want to infer the true y value for\footnote{DWH: Remember what this brain dump is for}. Leaving one out, and then estimating the prior and then inferring the posterior for each 1024 points, let alone over a million points is way too expensive. Luckily with large data sets, even at 1024, 1 datum is $\sim0.1\%$ of your dataset and therefore leaving that point out will not change the MML parameters of your denoised distribution, and with the \gaia\ dataset of over a million stars, we can safely step into the Empirical Bayes regime.

So using Bayes theorem
\begin{eqnarray}
p(y_{\true,n} \given y_n, \tilde{\sigma}_n) &=& p(y_n \given y_{\true,n}, \sigma_n)\,p(y_{\true,n}) \\
p(y_n \given y_{\true,n}, \sigma_n) &=& \mathcal{N}(y_n \given y_{\true,n}, \sigma_n) \\
p(y_{\true,n}) &=& \mathcal{N}(y_{\true,n} \given \hat{m}\, x_n + \hat{b}, \hat{t})
\label{eq:toyBayes}
\end{eqnarray}

Since everything is Gaussian, this can be done analytically by multiplying Gaussians. The blue points in Figure \ref{fig:line}, center panel, represent the posterior belief of the true value $p(y_{\true,n} \given y_n, \tilde{\sigma}_n)$. The black points represent the original, noisy data $y_n \pm \sigma_n$, the same as those highlighted in panel A, and the red points are the true values $y_{\true, n}$. Finally, Figure \ref{fig:line}, right panel, shows the full set of posterior beliefs for each $y_n$. The points are colored by their measurement variance $\sigma_n^2$ to show that noisier measurements are more influenced by the prior and get pulled closer to the center of the MML distribution, our prior, while more certain measurements remain closer to their measured values $y_n$.

LA: Conclusions and stuff about why this stuff is good
